---
title: "Introduction to mlr3tuningspaces"
description: |
  Apply predefined search spaces from scientific articles.
categories:
  - tuning
  - classification
author:
  - name: Marc Becker
date: 07-06-2021
aliases:
  - ../../../gallery/2021-07-06-introduction-to-mlr3tuningspaces/index.html
citation: true
---

{{< include ../../_setup.qmd >}}

# Scope

The package `r ref_pkg("mlr3tuningspaces")` offers a selection of published search spaces for many popular machine learning algorithms.
In this post, we show how to tune a `r ref("Learner", "mlr3 learners")` with these search spaces.

# Prerequisites

The packages `r ref_pkg("mlr3verse")` and `r ref_pkg("mlr3tuningspaces")` are required for this demonstration:

```{r 2021-07-06-introduction-to-mlr3tuningspaces-001, message = FALSE}
library(mlr3verse)
library(mlr3tuningspaces)
```

We initialize the random number generator with a fixed seed for reproducibility, and decrease the verbosity of the logger to keep the output clearly represented.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-002}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")
```

In the example, we use the `r ref("mlr_tasks_pima", text = "pima indian diabetes data set")` which is used to predict whether or not a patient has diabetes.
The patients are characterized by 8 numeric features, some of them have missing values.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-003}
# retrieve the task from mlr3
task = tsk("pima")

# generate a quick textual overview using the skimr package
skimr::skim(task$data())
```

# Tuning Search Space

For tuning, it is important to create a search space that defines the type and range of the hyperparameters.
A learner stores all information about its hyperparameters in the slot `$param_set`.
Usually, we have to chose a subset of hyperparameters we want to tune.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-004}
lrn("classif.rpart")$param_set
```

# Package

At the heart of `r ref_pkg("mlr3tuningspaces")` is the R6 class `r ref("TuningSpace")`.
It stores a list of `r ref("TuneToken")`, helper functions and additional meta information.
The list of `r ref("TuneToken")` can be directly applied to the `$values` slot of a learner's `r ref("ParamSet")`.
The search spaces are stored in the `r ref("mlr_tuning_spaces")` dictionary.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-005}
as.data.table(mlr_tuning_spaces)
```

We can use the sugar function `r ref("lts()")` to retrieve a `r ref("TuningSpace")`.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-006}
tuning_space_rpart = lts("classif.rpart.default")
tuning_space_rpart
```

The `$values` slot contains the list of of `r ref("TuneToken")`.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-007}
tuning_space_rpart$values
```

We apply the search space and tune the `r ref("Learner", "learner")`.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-008}
learner = lrn("classif.rpart")

learner$param_set$values = tuning_space_rpart$values

instance = tune(
  tuner = tnr("random_search"),
  task = tsk("pima"),
  learner = learner,
  resampling = rsmp ("holdout"),
  measure = msr("classif.ce"),
  term_evals = 10)

instance$result
```

We can also get the `r ref("Learner", "learner")` with search space already applied from the `r ref("TuningSpace")`.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-009}
learner = tuning_space_rpart$get_learner()
print(learner$param_set)
```

This method also allows to set constant parameters.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-010}
learner = tuning_space_rpart$get_learner(maxdepth = 15)
print(learner$param_set)
```

The `r ref("lts()")` function sets the default search space directly to a `r ref("Learner", "learner")`.

```{r 2021-07-06-introduction-to-mlr3tuningspaces-011}
learner = lts(lrn("classif.rpart", maxdepth = 15))
print(learner$param_set)
```
