---
title: "Most Popular Learners in mlr"
authors: ["Jabok Richter"]
date: 2017-03-30
categories: ["R", "r-bloggers"]
tags: ["learner", "algorithm", "popular", "machine-learning", "rstats"]
description: "An attempt to asses the popularity of mlr learners"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, eval = FALSE)
library(mlr)
library(stringi)
library(cranlogs)
library(data.table)
library(ggplot2)
```

For the development of [mlr](https://github.com/mlr-org/mlr) as well as for an "machine learning expert" it can be handy to know what are the most popular learners used.
Not necessarily to see, what are the top notch performing methods but to see what is used "out there" in the real world.
Thanks to the nice little package [cranlogs](https://github.com/metacran/cranlogs) from [metacran](https://www.r-pkg.org/) you can at least get a slight estimate as I will show in the following...

First we need to install the `cranlogs` package using `devtools`:

```{r install cranlogs, eval=FALSE}
devtools::install_github("metacran/cranlogs")
```

Now let's load all the packages we will need:

```{r libraries}
library(mlr)
library(stringi)
library(cranlogs)
library(data.table)
```

Do obtain a neat table of all available learners in _mlr_ we can call `listLearners()`.
This table also contains a column with the needed packages for each learner separated with a `,`.

```{r listLearners, warning=FALSE}
# obtain used packages for all learners
lrns = as.data.table(listLearners())
all.pkgs = stri_split(lrns$package, fixed = ",")
```

_Note:_ You might get some warnings here because you likely did not install all packages that _mlr_ suggests -- which is totally fine.

Now we can obtain the download counts from the _rstudio cran mirror_, i.e. from the last month.
We use `data.table` to easily sum up the download counts of each day.

```{r downloadCounts}
all.downloads = cran_downloads(packages = unique(unlist(all.pkgs)),
                               when = "last-month")
all.downloads = as.data.table(all.downloads)
monthly.downloads = all.downloads[, list(monthly = sum(count)), by = package]
```

As some learners need multiple packages we will use the download count of the package with the least downloads.

```{r aggregateDownloads}
lrn.downloads = sapply(all.pkgs, function(pkgs) {
  monthly.downloads[package %in% pkgs, min(monthly)]
})
```

Let's put these numbers in our table:

```{r firstTable, results='hide'}
lrns$downloads = lrn.downloads
lrns = lrns[order(downloads, decreasing = TRUE),]
lrns[, .(class, name, package, downloads)]
```

_Here are the first 5 rows of the table:_

```{r firstTableKnit, echo=FALSE, results='asis'}
knitr::kable(lrns[, .(class, name, package, downloads)][1:5])
```

Now let's get rid of the duplicates introduced by the distinction of the type _classif_, _regr_ and we already have our...

## Nearly final table

```{r finalTable, results='hide'}
lrns.small = lrns[, .SD[1,], by = .(name, package)]
lrns.small[, .(class, name, package, downloads)]
```

The top 20 according to the _rstudio cran mirror_:

```{r finalTableKnit, echo=FALSE, results='asis'}
knitr::kable(lrns.small[, .(class, name, package, downloads)][1:20])
```

As we are just looking for the packages let's compress the table a bit further and come to our...

## Final table

```{r compressTable, results='hide'}
lrns.pgks = lrns[,list(learners = paste(class, collapse = ",")),
                 by = .(package, downloads)]
lrns.pgks
```

_Here are the first 20 rows of the table:_

```{r compressTableKnit, echo=FALSE, results='asis'}
knitr::kable(lrns.pgks[1:20,])
```

And of course we want to have a small visualization:

```{r compressTablePlot}
library(ggplot2)
library(forcats)
lrns.pgks$learners = factor(lrns.pgks$learners, lrns.pgks$learners)
g = ggplot(lrns.pgks[20:1], aes(x = fct_inorder(stri_sub(
  paste0(package,": ",learners), 0, 64)), y = downloads, fill = downloads))
g + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("") +
  scale_fill_continuous(guide=FALSE)
```

## Remarks

This is not really representative of how popular each learner is, as some packages have multiple purposes (e.g. multiple learners).
Furthermore it would be great to have access to the [trending](https://www.r-pkg.org/trending) list.
Also [_most stars at GitHub_](https://www.r-pkg.org/starred) gives a better view of what the developers are interested in.
Looking for machine learning packages we see there e.g: [xgboost](https://github.com/dmlc/xgboost), [h2o](https://github.com/h2oai/h2o-3) and [tensorflow](https://github.com/rstudio/tensorflow).
