---
title: Deploy an mlr3 model with Plumber and Docker
categories:
  - mlr3pipelines
  - regression
author:
  - name: Lennart Schneider
    orcid: 0000-0003-4152-5308
  - name: Maximilian Mücke
    orcid: 0009-0000-9432-9795
date: 08-13-2020
description: |
  Deploy a trained mlr3 pipeline behind a Plumber API in a Docker image.
aliases:
  - ../../../gallery/2020-08-13-a-production-example-using-plumber-and-docker/index.html
---

{{< include ../../_setup.qmd >}}

A prediction service needs more than a fitted model. It also needs a stable input contract, an HTTP interface, and a portable runtime. This tutorial builds those pieces with `mlr3`, `plumber`, and Docker.

The example uses a small regression pipeline, but the deployment pattern also applies to other serializable `r ref("Learner", "Learner")` objects:

1. Train and serialize the complete learner.
2. Load it once when the API starts.
3. Convert each request into new data and call `$predict_newdata()`.
4. Build the training code, model artifact, API, and dependencies into one container image.

## Train and save the learner

We load the packages used for training and for the example prediction.

```{r index-001}
library(data.table)
library(mlr3)
library(mlr3pipelines)
```

The `r ref("mlr_tasks_boston_housing", "boston_housing")` task predicts the corrected median value of owner-occupied homes in thousands of US dollars. We retain one numeric, one integer, and one factor feature so that the API has to handle several feature types.

```{r index-002}
task_housing = tsk("boston_housing")
task_housing$select(c("crim", "tax", "town"))
```

The graph imputes missing numeric and factor values, aligns factor levels with those seen during training, and fits a regression tree.

```{r index-003}
graph = po("imputemedian") %>>%
  po("imputeoor") %>>%
  po("fixfactors") %>>%
  lrn("regr.rpart")

learner = as_learner(graph)
learner$train(task_housing)
```

A trained learner stores its feature names and data prototype. Its `$predict_newdata()` method uses this information to convert incoming columns to the types used during training. The separate feature metadata and custom conversion function required by the original version of this tutorial are therefore no longer necessary.

```{r index-004}
new_data = data.table(crim = 3.14, tax = 691L, town = "Newton")
learner$predict_newdata(new_data)$response
```

The complete learner can be serialized with `saveRDS()` and restored with `readRDS()`. The Docker build will run the following training script and create `model.rds` inside the image.

```{r index-005, eval=FALSE, code=readLines("mlr3_api/train_gl.R")}
```

This approach works for the regression tree in the example and for most learners. A learner whose model contains external pointers may first need `r ref("learner_marshal()")` before serialization and `r ref("learner_unmarshal()")` after loading.

## Define the Plumber API

The `r ref_pkg("plumber")` package turns annotated R functions into HTTP endpoints. The API loads `model.rds` once at startup and exposes two routes:

- `GET /health` confirms that the process started and loaded the model.
- `POST /predict` accepts an `observations` array and returns one prediction per row.

The prediction route converts the parsed observations to a `r ref("data.table::data.table", "data.table")` and passes them to `$predict_newdata()`. The learner checks the features, converts compatible types, and applies the fitted pipeline.

```{r index-006, eval=FALSE, code=readLines("mlr3_api/predict_gl.R")}
```

The request body uses a named array rather than a bare JSON array. This gives the endpoint a stable top-level contract that can be extended later with fields such as a model version or request identifier.

The example relies on Plumber's default error handling. A public API should validate its request schema and return structured client errors.

## Build the Docker image

The [`rstudio/plumber`](https://hub.docker.com/r/rstudio/plumber) image provides R, `plumber`, `pak`, and an entrypoint that serves a Plumber file on port 8000. The Dockerfile installs `mlr3` and `mlr3pipelines` with `pak`, copies the training and API scripts, and trains the learner.

The completed image contains the trained learner and starts the API on port 8000.

```{r index-007, engine="bash", eval=FALSE, code=readLines("mlr3_api/Dockerfile")}
```

From the `mlr3_api` directory, build and start the image:

```{r index-008, engine="bash", eval=FALSE}
docker pull rstudio/plumber
docker build --tag mlr3-plumber-demo .
docker run --detach --rm \
  --name mlr3-api \
  --publish 8000:8000 \
  mlr3-plumber-demo
```

Check the health endpoint and submit two observations:

```{r index-009, engine="bash", eval=FALSE}
curl http://127.0.0.1:8000/health

curl \
  --request POST \
  --header "Content-Type: application/json" \
  --data '{"observations":[
    {"crim":3.14,"tax":691,"town":"Newton"},
    {"crim":0.03,"tax":222,"town":"Cambridge"}
  ]}' \
  http://127.0.0.1:8000/predict
```

The response contains a numeric vector in the same row order as the request.

```json
{"predictions":[33.3011,23.9582]}
```

Stop the named container when finished:

```{r index-010, engine="bash", eval=FALSE}
docker stop mlr3-api
```

Rebuilding the image retrains the model. In an automated deployment, the same build should also run data checks and model evaluation, and it should only publish the image if those checks pass.

::: {.callout-warning}
The container provides a portable model service, not a complete production perimeter. Pin the base image and R package versions when builds must be reproducible. Put authentication, TLS termination, request-size limits, rate limits, and audit logging in front of the API. Run multiple replicas or a suitable process manager for concurrency, and do not expose this demonstration directly to the internet.
:::

{{< include ../../_session_info.qmd >}}
