---
title: "Announcing mlr3spatial"
description: |
  We are happy to announce that mlr3spatial has been released on CRAN in November 2021.
author:
  - name: Marc Becker
    url: https://github.com/be-marc
  - name: Patrick Schratz
    url: https://github.com/pat-s
date: 2021-12-01
categories:
    - R
    - CRAN
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

We are happy to announce that [mlr3spatial](https://mlr3spatial.mlr-org.com/) has been released on [CRAN](https://cran.r-project.org/web/packages/mlr3spatial/index.html) in November 2021.
[mlr3spatial](https://mlr3spatial.mlr-org.com/) simplifies the handling of spatial objects in the [mlr3](https://github.com/mlr-org) ecosystem.
Before [mlr3spatial](https://mlr3spatial.mlr-org.com/), the user had to extract tabular data from spatial objects to train a model or predict spatial data.

Now, mlr3 [Tasks](https://mlr3.mlr-org.com/reference/Task.html) can directly read from spatial objects via specialized [Data Backends](https://mlr3.mlr-org.com/reference/DataBackend.html).
Such tasks can be used to train a model or to perform resampling just like any other `mlr3` task.
We support spatial raster objects created by the [terra](https://CRAN.R-project.org/package=terra), [raster](https://CRAN.R-project.org/package=raster) and [stars](https://CRAN.R-project.org/package=stars) packages with [`DataBackendRaster`](https://mlr3spatial.mlr-org.com/reference/DataBackendRaster.html).
Additionally, vector data created with the [sf](https://CRAN.R-project.org/package=sf) package is handled with [`DataBackendVector`](https://mlr3spatial.mlr-org.com/reference/DataBackendVector.html).

The `predict_raster()` function creates spatial rasters and features with predictions from mlr3 learners.
We only have to pass a task with a spatial data backend which provides the data and spatial reference.
To avoid memory issues with large raster files, prediction is done in chunks.
For this, the raster map is divided into multiple horizontal strips.
The vertical extension of these strips is controlled by the `chunksize` parameter.
The actual memory usage per core is a multiple of the specified chunk size.
We choose a default chunk size of 200 Megabytes which should work on most consumer grade machines.
If more memory is available, a larger chunk size accelerates the prediction process.

One after the other, the raster chunks are loaded into memory and the prediction is written to disk.
Finally, the complete raster is available on disk.
The learner can also make use of future-based parallelization to accelerate the predictions.
The vignette on [“Benchmarking parallel predictions”](https://mlr3spatial.mlr-org.com/articles/benchmark.html) showcases the parallelization capabilities of [mlr3spatial](https://mlr3spatial.mlr-org.com/).

## Use Case - Landsat7 data as {stars} object

### Data Preparation

```{r mlr3spatial-001, include=FALSE}
library("stars")
```

```{r mlr3spatial-002}
library("mlr3")
library("mlr3spatial")
```

First, the TIFF files is read via `stars::read_stars()` and put into a `DataBackendRaster`.
The DataBackend is then used to create a regression task with the response being `layer.1`.

```{r mlr3spatial-003, error=TRUE}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
stack = stars::read_stars(tif)

backend = as_data_backend(stack)
task = as_task_regr(backend, target = "band1")

print(task)
```

For large raster files with millions of values it helps to predict in parallel.
To enable this, set `learner$parallel_predict = TRUE` and initiate a parallel plan via {future}, e.g. via `plan("multisession")`.
Since this is only an example, parallelization is not enabled here.
Here we will use a simple regression tree as an example learner.
In practice you might want to use a different learner - you can find an overview of available learners [here](https://mlr3extralearners.mlr-org.com/articles/learners/list_learners.html).

```{r mlr3spatial-004}
learner = lrn("regr.rpart")
set.seed(42)
row_ids = sample(1:task$nrow, 500)
learner$train(task, row_ids = row_ids)

print(learner)
```

### Prediction

For prediction `predict_spatial()` is used.
It will return a raster file which contains the predictions.
Users can select which R spatial format the returned raster should have.
Note that by default `mlr3spatial` operates with `SpatRaster` objects from package `terra`.
If a different output format is requested (e.g. `"stars"`), coercion is happening in the background which might take some time.

```{r mlr3spatial-005, warning=FALSE}
ras = predict_spatial(task, learner, format = "stars")
names(ras) = "cadmium"

print(ras)
```

### Visualization

Finally we can plot the predictions.
The color vector is extracted from the "viridis" color palette via `dput(viridis::viridis_pal()(5))` and provided to the S3 `plot()` call, which makes use of the S3 plot method within the `stars` package in this scenario.

```{r mlr3spatial-006}
plot(ras, col = c("#440154FF", "#443A83FF", "#31688EFF", "#21908CFF", "#35B779FF", "#8FD744FF", "#FDE725FF"))
```



