---
title: "Asynchronous Optimization Runtime Benchmarks"
sidebar: false
toc: true
cache: false
lazy-cache: false
freeze: true
format:
  html:
    fig-width: 12
    fig-height: 9
bibliography: ../bibliography.bib
---

{{< include ../_setup.qmd >}}

```{r}
#| include: false
library(ggplot2)
library(data.table)
library(mlr3misc)
library(gt)

data = fread(here::here("mlr-org/benchmarks/xgboost_1280_30m.csv"))
setnames(data, new = gsub("xgboost\\.", "", names(data)))
setorderv(data, "holdout_auc")
cols_x = c("alpha", "colsample_bylevel", "colsample_bytree", "eta", "lambda", "max_depth", "subsample", "nrounds")

data_hyperband = data[c("asha", "asha_hotstart"), , on = "tuner"][,
  list(
    mean_auc = mean(holdout_auc),
    max_auc = max(holdout_auc),
    mean_runtime = mean(runtime_learners),
    n_configs = .N),
  by = c("tuner", "stage")
]

data_rank = map_dtr(seq(10, 1800, by = 30), function(i) {
  data_i = data[walltime < i]
  setkey(data_i, walltime)
  data_i = data_i[, tail(.SD, 1), by = tuner]
  data_i[, rank := frank(incumbent, ties.method = "min")]
})
```

# Scope

This report examines the runtime performance of asynchronous optimization algorithms implemented in the mlr3tuning.

The algorithms optimize 8 hyperparameters of the XGBoost learner on the Bank Marketing dataset.
This dataset contains client data from a Portuguese bank's marketing campaign, including demographics, financial information, and previous campaign interactions, used to predict term deposit subscriptions.
To ensure an unbiased evaluation of the configurations, a separate validation set was reserved.
The remaining data was subjected to 3-fold cross-validation during the optimization process.
The performance of each configuration was evaluated using the area under the receiver operating characteristic curve (AUC).

The experiments were conducted on 10 nodes, each equipped with 128 cores, totaling 1,280 workers.
The optimization process ran for 30 minutes, amounting to 640 CPU hours.

# Optimization Algorithms

We compare four asynchronous optimization algorithms.
Asynchronous Distributed Bayesian Optimization (ADBO) [@egele_2023] runs sequential [Bayesian optimization](https://mlr3book.mlr-org.com/chapters/chapter5/advanced_tuning_methods_and_black_box_optimization.html#sec-bayesian-optimization) on multiple workers in parallel.
Each worker maintains its own surrogate model (a random forest) and selects the next hyperparameter configuration by maximizing an upper confidence bounds acquisition function.
To promote a varying exploration-exploitation tradeoff on the workers, the acquisition functions are initialized with different lambda values ranging from 0.1 to 10.
When a worker completes an evaluation, it asynchronously sends the result to its peers via a Redis data base; each worker then updates its local model with this shared information.
This decentralized design enables workers to proceed independently — eliminating the need for a central coordinator that could become a bottleneck in large-scale optimization scenarios.
The number of boosting rounds is separately optimized using [early stopping](https://mlr3book.mlr-org.com/chapters/chapter15/predsets_valid_inttune.html#sec-internal-tuning) on the test set.
To start the optimization process with a diverse set of configurations, a Sobol sample with 2,000 configurations is generated.

Asynchronous Successive Halving Algorithm (ASHA) [@li_2020] belongs to the family of [hyperband](https://mlr3book.mlr-org.com/chapters/chapter5/advanced_tuning_methods_and_black_box_optimization.html#sec-hyperband) algorithms.
The algorithm is an extension Successive Halving Algorithm, which initially allocates minimal resources to many configurations and then progressively increases resource allocation only for those configurations that perform well.
One of the key innovations of ASHA is its asynchronous approach to promotions.
Unlike traditional methods that require all configurations in a given stage to complete before moving on, ASHA promotes promising configurations as soon as they meet the performance criteria.
This strategy eliminates the bottleneck caused by slower or underperforming models, ensuring that the overall process remains efficient and that computing resources are used effectively.
We setup the algorithm to allocate 9 boosting rounds in the first stages and increases the number of boosting rounds by a factor of 3 in each subsequent stage until the maximum number of boosting rounds of 2187 is reached.
The hotstart variant of ASHA continues training from the previous stage rather than retraining from scratch.

As a baseline, we implement parallel random search [@bergstra_2012] across all workers.
This method samples hyperparameter configurations uniformly at random from the search space, and the optimal number of boosting rounds is determined using early stopping on the test set.

The code shows a simplified version of the benchmark experiment.
We removed parts that are specific to the high performance cluster.

```{r}
#| eval: false
library(rush)
library(mlr3tuning)
library(mlr3learners)
library(mlr3oml)
library(mlr3pipelines)
library(mlr3hyperband)

set.seed(4356)

config = redux::redis_config(
  host = "10.14.20.237",
  port = "6362"
)

rush_plan(n_workers = 1280, config = config, worker_type = "script")

learner = as_learner(po("encodeimpact") %>>% lrn("classif.xgboost",
  eta               = to_tune(1e-4, 1, logscale = TRUE),
  max_depth         = to_tune(1, 20),
  colsample_bytree  = to_tune(1e-1, 1),
  colsample_bylevel = to_tune(1e-1, 1),
  lambda            = to_tune(1e-3, 1e3, logscale = TRUE),
  alpha             = to_tune(1e-3, 1e3, logscale = TRUE),
  subsample         = to_tune(1e-1, 1),
  nrounds           = to_tune(p_int(9, 2187, tags = "budget")),
  predict_type = "prob",
  id = "xgboost"))

# download task
otask = otsk(359982)
task = as_task(otask)

# split task into training-test and validation set
splits = partition(task, ratio = 0.8)
task_holdout = task$clone()
task_holdout$filter(splits$test)
task$filter(splits$train)

# create tuning instance
instance = ti_async(
  task = task,
  learner = learner,
  resampling = rsmp("cv", folds = 3),
  measures = msr("classif.auc"),
  terminator = trm("run_time", secs = 1800L),
  store_benchmark_result = FALSE,
  store_models = FALSE,
  callbacks = clbk("mlr3tuning.holdout", task = task_holdout)
)

# selected tuner
tuner = tnr("async_successive_halving", eta = 3)

# optimize
tuner$optimize(instance)
```

# Results

The following plots show the performance of the algorithms over time.
The performance differences are small and all algorithms quickly converge to good configurations.
The ASHA algorithm with hotstarting outperforms the ADBO and Random Search algorithms in terms of validation AUC.

```{r}
#| echo: false
#| warning: false
#| fig-cap: Performance over time for each optimizer.
#| fig-subcap:
#|   - "Validation AUC Performance"
#|   - "Rank Performance"
#| layout-ncol: 1
ggplot(data, aes(x = walltime, y = incumbent, group = tuner)) +
  geom_line(aes(color = tuner)) +
  scale_y_continuous(limits = c(0.93, 0.94)) +
  scale_color_discrete(
    name = "Tuner",
    labels = c("ADBO", "ASHA", "ASHA Hotstart", "Random Search")) +
  xlab("Walltime (s)") +
  ylab("Validation AUC") +
  theme_minimal()

ggplot(data_rank, aes(x = walltime, y = rank, group = tuner)) +
  geom_line(aes(color = tuner)) +
  scale_color_discrete(
    name = "Tuner",
    labels = c("ADBO", "ASHA", "ASHA Hotstart", "Random Search")) +
  xlab("Walltime (s)") +
  ylab("Rank") +
  theme_minimal()
```

The next table shows the final configurations selected by the optimizers and their validation AUC.
Compared to the other algorithms, random search selected a configuration with a higher number of boosting rounds.

```{r}
#| tbl-cap: Final configurations selected by the optimizers.
#| echo: false
#| column: page
gt(data[, tail(.SD, 1), by = tuner][, c("tuner", "holdout_auc", "runtime_learners", cols_x), with = FALSE]) %>%
  fmt_number(
    columns = c("runtime_learners", "alpha", "colsample_bylevel", "colsample_bytree", "eta", "lambda", "subsample"),
    decimals = 2
  ) %>%
  fmt_number(
    columns = c("holdout_auc"),
    decimals = 4
  ) %>%
  cols_label(
    tuner = "Tuner",
    holdout_auc = "Validation AUC",
  )
```

The ASHA algorithms evaluated 12 to 22 times more configurations than ADBO and Random Search due to the low fidelity evaluations in the first stages.


```{r}
#| fig-cap: Number of configurations evaluated by each optimizer.
#| echo: false
ggplot(data[, list(n_configs = .N), by = tuner], aes(x = tuner, y = n_configs, group = tuner)) +
  geom_bar(aes(fill = tuner), stat = "identity") +
  labs(x = "Optimizer",
       y = "Number of Configurations") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("ADBO", "ASHA", "ASHA Hotstart", "Random Search")) +
  theme_minimal()
```

Figure @fig-hyperband shows the number of evaluated hyperparameter configurations per stage for ASHA and ASHA Hotstart.
The number of configurations is reduced in each stage by a factor of 3.
With hotstarting more configurations are evaluated in all stages compared to the base ASHA algorithm.
The number of evaluated configurations almost doubles due to the reduced training time of the models.

```{r}
#| fig-cap: Number of evaluated hyperparameter configurations per stage for ASHA and ASHA Hotstart.
#| echo: false
#| label: fig-hyperband
ggplot(data_hyperband, aes(x = stage, y = n_configs, group = tuner)) +
  geom_bar(aes(fill = tuner), stat = "identity", position = "dodge") +
  labs(title = "Number of Evaluations per Stage",
       x = "Stage",
       y = "Number of Evaluations",
       fill = "Tuner") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("ASHA", "ASHA Hotstart"), end = 0.8) +
  theme_minimal()
```

As an example, we show the marginal plot of the learning rate `eta` and regularization parameter `alpha` from the ASHA Hotstart run.
The plot shows the configurations evaluated by the algorithm in each stage.
Each point represents a configuration, and the color shows the validation AUC value.
In the fist stage many configurations are tried quickly with a small training budget.
We can see that large `alpha` values produce lower AUC values and are not promoted to the next stage.
As the algorithm progresses, it focuses on more promising regions of the hyperparameter space.
By the last stages we have fewer configurations but a large number of boosting rounds.
The algorithm favors configurations in the middle range of `eta` values.

```{r}
#| echo: false
#| fig-cap: |
#|   Marginal plot of `eta` and `alpha` for ASHA Hotstart.
#|   Each point represents a configuration evaluated by ASHA Hotstart.
#|   The color shows the validation AUC value.
#|   The facets show the number of boosting rounds or stages.
ggplot(data[tuner == "asha_hotstart"], aes(x = eta, y = alpha, color = holdout_auc)) +
  geom_point(size = 1/3) +
  xlab("eta") +
  ylab("alpha") +
  facet_wrap(~nrounds) +
  scale_color_viridis_c(name = "Validation AUC") +
  theme_minimal()
```

One of the main reasons to develop asynchronous algorithms for mlr3tuning was the low cpu utilization of the batch parallelization approach.
The following table shows the mean utilization of the workers.
The utilization is calculated as the ratio of the runtime of the learners to the runtime of the workers.
All algorithms show an overhead around 7% due to the overhead of `resample()` and the communication between the workers.

```{r}
#| tbl-cap: Mean utilization of the workers for each optimizer.
#| echo: false
aggr = data[, list(
  runtime_learners = sum(runtime_learners),
  runtime_worker = as.numeric(difftime(max(timestamp_ys), min(timestamp_xs), units = "secs"))
  ), by = c("worker_id", "tuner")]
aggr[, utilization := runtime_learners / runtime_worker]

gt(aggr[, .(mean_utilization = mean(utilization)), by = tuner]) %>%
  fmt_percent(columns = "mean_utilization", decimals = 1) %>%
  cols_label(
    tuner = "Tuner",
    mean_utilization = "Mean Utilization"
  )
```

