---
sidebar: false
bibliography: publications.bib
title: "Ecosystem"
---

{{< include _setup.qmd >}}

The mlr3 ecosystem is a collection of R packages for machine learning.
The base package mlr3 only provides the basic building blocks for machine learning.
The extensions packages extent mlr3 with functionality for additional task types, learning algorithms, tuning algorithms, feature selection strategies, visualizations or preprocessing capabilities.
The packages are listed bellow with a short description.
For more information about the packages, check out their respective homepages.

The dot next to the package name indicates the [lifecycle stage](https://lifecycle.r-lib.org/articles/stages.html).

- Packages with a green dot <i class="bi-circle-fill status-stable" title = "[status-stable]"></i> are stable.
- Experimental packages are marked with an orange dot <i class="bi-circle-fill status-experimental" title = "[status-experimental]"></i>.
- Planned packages are marked with a red dot <i class="bi-circle-fill status-planned" title = "[status-planned]"></i>.

![Graph of Extension Packages](https://raw.githubusercontent.com/mlr-org/mlr3/master/man/figures/mlr3verse.svg?sanitize=true)

If you use our packages in your research, please cite our articles on mlr3 [@mlr3], mlr3proba [@mlr3proba] or mlr3pipelines [@mlr3pipelines].
To get the citation information of other packages, call
```{r eval = FALSE, echo = TRUE}
citation("[package]")
```

```{r}
#| include: false
package_grid = function(content) {
  div(class = "grid grid-three",
    pmap(content, function(category, name, status, url, text, on_cran, ...) {
        has_homepage = grepl("mlr-org.com", url, ignore.case = TRUE)
        div(class = "grid__item",
          h3(class = "grid__title",
            a(class = "grid__title-link", href = url, name),
            tags$i(
              class = sprintf("bi-circle-fill grid__status grid__status--%s", status),
              title = sprintf("[status-%s]", status)
            )
          ),
          p(class = "grid__text--footer", paste0(text, ".")),
          div(class = "grid__item-footer",
            if (has_homepage) a(class = "grid__link-footer", href = url, "aria-label" = "Go to homepage", tags$i(class = "bi bi-house grid__icon-footer", title = "Homepage")),
            if (has_homepage) a(class = "grid__link-footer", href = paste0(url, "reference/"), "aria-label" = "Go to Reference Manual", tags$i(class = "bi bi-question-circle grid__icon-footer", title = "Reference Manual")),
            a(class = "grid__link-footer", href = sprintf("https://github.com/mlr-org/%s", name), "aria-label" = "Go to Project Page", tags$i(class = "bi bi-github grid__icon-footer", title = "Github Project")),
            if (on_cran) a(class = "grid__link-footer", href = sprintf("https://cran.r-project.org/package=%s", name), "aria-label" = "Go to CRAN page", tags$i(class = "fab fa-r-project grid__icon-footer", title = "CRAN Website"))
          )
        )
    })
  )
}

# Text description should be one sentence
content = rowwise_table(
    ~category,       ~name,               ~status,        ~url,                                         ~text,
    "core",          "mlr3",              "stable",       "https://mlr3.mlr-org.com/",                  "Basic building blocks for machine learning",
    "core",          "mlr3verse",         "stable",       "https://mlr3verse.mlr-org.com/",             "Meta-package intended to simplify both installation and loading of packages from the mlr3 ecosystem",
    "optimization",  "mlr3tuning",        "stable",       "https://mlr3tuning.mlr-org.com/",            "Hyperparameter tuning for mlr3 learners",
    "optimization",  "mlr3tuningspaces",  "stable",       "https://mlr3tuningspaces.mlr-org.com/",      "Collection of search spaces for hyperparameter tuning",
    "optimization",  "mlr3hyperband",     "stable",       "https://mlr3hyperband.mlr-org.com/",         "Successive halving and hyperband tuner for mlr3tuning",
    "optimization",  "mlr3mbo",           "experimental", "https://mlr3mbo.mlr-org.com/",               "Model-based optimization for mlr3tuning",
    "optimization",  "miesmuschel",       "experimental", "https://github.com/mlr-org/miesmuschel/",    "Flexible mixed integer evolutionary strategies",
    "optimization",  "mlr3automl",        "experimental", "https://github.com/mlr-org/mlr3automl/",     "Automated machine learning",
    "featsel",       "mlr3filters",       "stable",       "https://mlr3filters.mlr-org.com/",           "Filter Feature Selection",
    "featsel",       "mlr3fselect",       "stable",       "https://mlr3fselect.mlr-org.com/",           "Wrapper Feature Selection",
    "data",          "mlr3db",            "stable",       "https://mlr3db.mlr-org.com/",                "Data backend to transparently work with databases",
    "data",          "mlr3oml",           "stable",       "https://mlr3oml.mlr-org.com/",               "Connector to OpenML",
    "data",          "mlr3data",          "stable",       "https://mlr3data.mlr-org.com/",              "Data sets and tasks",
    "learners",      "mlr3learners",      "stable",       "https://mlr3learners.mlr-org.com/",          "Essential learners for mlr3, maintained by the mlr-org team",
    "learners",      "mlr3extralearners", "stable",       "https://mlr3extralearners.mlr-org.com/",     "Extra learners for mlr3, implemented by the community",
    "learners",      "mlr3keras",         "experimental", "https://github.com/mlr-org/mlr3keras/",      "Deep learning with Keras",
    "visualization", "mlr3viz",           "stable",       "https://mlr3viz.mlr-org.com/",               "Visualizations for tasks, predictions, resample results and benchmarks",
    "pipelines",     "mlr3pipelines",     "stable",       "https://mlr3pipelines.mlr-org.com/",         "Dataflow programming toolkit",
    "tasks",         "mlr3spatiotempcv",  "stable",       "https://mlr3spatiotempcv.mlr-org.com/",      "Spatiotemporal resampling and visualization methods",
    "tasks",         "mlr3cluster",       "stable",       "https://mlr3cluster.mlr-org.com/",           "Cluster analysis",
    "tasks",         "mlr3proba",         "stable",       "https://mlr3proba.mlr-org.com/",             "Probabilistic predictions",
    "tasks",         "mlr3ordinal",       "experimental", "https://mlr3ordinal.mlr-org.com/",           "Ordinal regression",
    "tasks",         "mlr3multioutput",   "experimental", "https://github.com/mlr-org/mlr3multioutput", "Storing and working with multi-output tasks",
    "tasks",         "mlr3temporal",      "experimental", "https://mlr3temporal.mlr-org.com/",          "Time series analysis",
    "tasks",         "mlr3spatial",       "stable",       "https://mlr3spatial.mlr-org.com/",           "Spatial data backends and prediction functions",
    "tasks",         "mlr3multilabel",    "planned",      "https://github.com/mlr-org/mlr3multilabel",  "Multi-Label Classification",
    "tasks",         "mlr3fda",           "planned",      "https://github.com/mlr-org/mlr3fda",         "Functional Data Analysis",
    "utils",         "bbotk",             "stable",       "https://bbotk.mlr-org.com/",                 "Black-box optimization toolkit",
    "utils",         "paradox",           "stable",       "https://paradox.mlr-org.com/",               "Universal parameter space description and tools",
    "utils",         "mlr3misc",          "stable",       "https://mlr3misc.mlr-org.com/",              "Miscellaneous helper functions for mlr3",
    "utils",         "mlr3measures",      "stable",       "https://mlr3measures.mlr-org.com/",          "Performance measures for supervised learning",
    "utils",         "mlr3benchmark",     "stable",       "https://mlr3benchmark.mlr-org.com/",         "Analysis and tools for benchmarking",
    "utils",         "mlr3fairness",      "experimental", "https://mlr3fairness.mlr-org.com/",          "Fairness in Machine Learning",
    "utils",         "mlr3batchmark",     "experimental", "https://github.com/mlr-org/mlr3batchmark/",  "Connector between mlr3 and batchtools"
)
# add CRAN link
content[, on_cran := name %in% rownames(available.packages())]
# index by categoty
# sort by status: stable first, then experimental, then planned
content[, status := factor(status, levels = c("stable", "experimental", "planned"), ordered = TRUE)]
# sort by status, keep order of table
setkeyv(content, c("category", "status"))
# sort by name
# setkeyv(content, c("category", "name"))
```

## <i class="bi-house-heart"></i> Core

```{r}
#| echo: false
package_grid(content[.("core")])
```


## <i class="bi-sliders"></i> Optimization

```{r}
#| echo: false
package_grid(content[.("optimization")])
```

## <i class="bi-funnel"></i> Feature Selection

```{r}
#| echo: false
package_grid(content[.("featsel")])
```

## <i class="bi-hdd"></i> Data

```{r}
#| echo: false
package_grid(content[.("data")])
```

## <i class="bi-pen"></i> Learners

```{r}
#| echo: false
package_grid(content[.("learners")])
```

## <i class="bi-bar-chart-line"></i> Visualization

```{r}
#| echo: false
package_grid(content[.("visualization")])
```

## <i class="bi-share"></i> Pipelines

```{r}
#| echo: false
package_grid(content[.("pipelines")])
```

## <i class="bi-list-task"></i> Tasks

```{r}
#| echo: false
package_grid(content[.("tasks")])
```

## <i class="bi-balloon"></i> Utilities

```{r}
#| echo: false
package_grid(content[.("utils")])
```
