---
sidebar: false
toc: false
---

# Feature Selection Wrapper

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(mlr3website)
library(htmltools)
library(mlr3misc)
library(data.table)
library(mlr3fselect)
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")

content = as.data.table(mlr_fselectors, objects = TRUE)[, .(key, label, packages, properties, object)]

# add links
content[, key := link_keys(key, prefix = "mlr_fselectors")]
content[, packages := link_pkgs(packages)]

# remove content
content[, properties := map(properties, function(x) setdiff(x, "dependencies"))]

# add line breaks to cells
cols = c("properties", "packages")
content[, (cols) := map(.SD, function(col) map_chr(col, function(x) paste0(x, collapse = "<br/>"))), .SDcols = cols]

content[, object := NULL]
setnames(content, c("Key", "Label", "Packages", "Properties"))
```

Feature selection wrappers can be found in the `r mlr_pkg("mlr3fselect")` packages.
The goal is to find the best subset of features with respect to a `r ref("Measure", text = "performance measure")` in an iterative fashion.

```{r}
#| column: page
DT::datatable(
  content,
  style = "bootstrap",
  escape = FALSE,
  options = list(paging = FALSE, scrollX = TRUE, info = FALSE),
  rownames = FALSE)
```

## Example Usage

Run a `r ref("mlr_fselectors_sequential", text = "sequential feature selection")` on the `r ref("mlr_tasks_pima", text = "Pima Indian Diabetes")` data set.

```{r, echo = TRUE}
library(mlr3verse)

# retrieve task
task = tsk("pima")

# load learner
learner = lrn("classif.rpart")

# feature selection on the pima indians diabetes data set
instance = fselect(
  method = fs("sequential"),
  task = task,
  learner = learner,
  resampling = rsmp("holdout"),
  measure = msr("classif.ce")
)

# best performing feature subset
instance$result

# subset the task and fit the final model
task$select(instance$result_feature_set)
learner$train(task)

print(learner)
```
