---
title: "Machine Learning in R"
description: |
  Website of the mlr3 ecosystem
preview: images/logo_color.png
site: distill::distill_website
output:
    distill::distill_article:
      includes:
        before_body: _banner.html
---

An open-source machine learning framework providing a unified interface for
machine learning in the R language.

```{r, echo = FALSE}
# Feature grid
library(mlr3misc)
library(htmltools)
library(data.table)

content = rbindlist(list(
  list(icon = "code-square", text = "A well designed and easy to learn interface."),
  list(icon = "pen", text = "More than 100 connected machine learning algorithms."),
  list(icon = "box", text = "Light on dependencies."),
  list(icon = "stack", text = "Quick parallelization with the `future` package."),
  list(icon = "sliders", text = "State-of-the-art optimization algorithms."),
  list(icon = "share", text = "Dataflow programming with pipelines.")
))

div(class = "grid-container",
  pmap(content, function(icon, text) {
    div(class = "grid-item",
      div(class="grid-icon",
        tags$i(class = sprintf("bi-%s", icon))
      ),
      text
    )
  })
)
```

```{r, echo = FALSE}
# Social buttons
content = rbindlist(list(
  list(name = "GitHub", url = "https://github.com/mlr-org/mlr3", icon = "github"),
  list(name = "Mattermost", url = "https://lmmisld-lmu-stats-slds.srv.mwn.de/mlr_invite/", icon = "chat-dots"),
  list(name = "Twitter", url = "https://twitter.com/bbischl", icon = "twitter"),
  list(name = "StackOverflow", url = "https://stackoverflow.com/questions/tagged/mlr3", icon = "stack-overflow")
))

div(class = "btn-container",
  pmap(content, function(name, url, icon) {
    a(href = url, class="btn",
      tags$i(class = sprintf("bi-%s", icon)),
      name
    )
  })
)
```

## Quick Start

The mlr3verse meta-package installs mlr3 and some of the most important extension packages:

```{r, eval=FALSE}
install.packages("mlr3verse")
```

### Basic Example

```{r, include=FALSE}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")
```

Simple resampling:

1. Load task and learner. Optionally, set hyperparameter values.
1. Choose resampling strategy.
1. Run automatic resampling
1. Score with classification accuracy

```{r, eval=FALSE}
library(mlr3verse)

task = tsk("penguins")
learner = lrn("classif.rpart", cp = .01)

resampling = rsmp("cv", folds = 3L)

rr = resample(task, learner, resampling)

rr$score(msr("classif.acc"))
```

### Tuning Example

Quick tuning of a simple classification tree on the on penguins data set:

1. Choose task.
1. Load learner and set search space in one go.
1. Run hyperparameter tuning with random search.
1. Get best performing hyperparameter configuration.

```{r, eval=FALSE}
library(mlr3tuningspaces)

task = tsk("penguins")

learner = lts(lrn("classif.rpart"))

instance = tune(
  method = "random_search",
  task = task,
  learner = learner,
  resampling = rsmp("cv", folds = 3),
  measure = msr("classif.acc"),
  term_evals = 10,
  batch_size = 5
)

instance$result
```

### Pipelines Example

Create a stacked learner:

1. Load task.
1. Choose list of base learners and super learner.
1. Convert to graph learner.
1. Train on german credit task.

```{r, eval=FALSE}
library(mlr3verse)
library(mlr3pipelines)

task = tsk("german_credit")

base_learners = list(
  lrn("classif.rpart", predict_type = "prob"),
  lrn("classif.kknn", predict_type = "prob")
)
super_learner = lrn("classif.log_reg")

graph_stack = pipeline_stacking(base_learners, super_learner)
graph_learner = as_learner(graph_stack)

graph_learner$train(task)
```

## Citing mlr3

If you use `mlr3`, please cite our JOSS article:

<div class="citation">
```
@Article{mlr3,
  title = {{mlr3}: A modern object-oriented machine learning framework in {R}},
  author = {Michel Lang and Martin Binder and Jakob Richter and Patrick Schratz and Florian Pfisterer and Stefan Coors and Quay Au and Giuseppe Casalicchio and Lars Kotthoff and Bernd Bischl},
  journal = {Journal of Open Source Software},
  year = {2019},
  month = {dec},
  doi = {10.21105/joss.01903},
  url = {https://joss.theoj.org/papers/10.21105/joss.01903},
}
```
</div>
