---
title: "Machine Learning in R"
description: |
  Website of the mlr3 ecosystem
preview: images/logo_color.png
site: distill::distill_website
output:
    distill::distill_article:
      includes:
        before_body: _banner.html
---

An open-source machine learning framework providing a unified interface for
machine learning in the R language.

<div class="grid-container">
  <div class="grid-item">
  <div class="grid-icon"><i class="bi-code-square"></i></div>
  A well designed and easy to learn interface.
  </div>
  <div class="grid-item">
  <div class="grid-icon"><i class="bi-pen"></i></div>
  More than 100 connected machine learning algorithms.
  </div>
  <div class="grid-item">
  <div class="grid-icon"><i class="bi-box"></i></div>
  Light on dependencies.
  </div>
  <div class="grid-item">
  <div class="grid-icon"><i class="bi-stack"></i></div>
  Quick parallelization with the `future` package.
  </div>
  <div class="grid-item">
  <div class="grid-icon"><i class="bi-sliders"></i></div>
  State-of-the-art optimization algorithms.
  </div>
  <div class="grid-item">
  <div class="grid-icon"><i class="bi-share"></i></div>
  Dataflow programming with pipelines.</div>
</div>

<div class="btn-container">
<a href="https://github.com/mlr-org/mlr3" class="btn"><i class="bi-github"></i> GitHub</a>
<a href="https://lmmisld-lmu-stats-slds.srv.mwn.de/mlr_invite/" class="btn"><i class="bi-chat-dots"></i> Mattermost</a>
<a href="https://twitter.com/bbischl" class="btn"><i class="bi-twitter"></i> Twitter</a>
<a href="https://stackoverflow.com/questions/tagged/mlr3" class="btn"><i class="bi-stack-overflow"></i> StackOverflow</a>
</div>

## Quick Start

The mlr3verse meta-package installs mlr3 and some of the most important extension packages:

```{r, eval=FALSE}
install.packages("mlr3verse")
```

### Basic Example

```{r, include=FALSE}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")
```

Simple resampling:

1. Load task and learner. Optionally, set hyperparameter values.
1. Choose resampling strategy.
1. Run automatic resampling
1. Score with classification accuracy

```{r, eval=FALSE}
library(mlr3verse)

task = tsk("penguins")
learner = lrn("classif.rpart", cp = .01)

resampling = rsmp("cv", folds = 3L)

rr = resample(task, learner, resampling)

rr$score(msr("classif.acc"))
```

### Tuning Example

Quick tuning of a simple classification tree on the on penguins data set:

1. Choose task.
1. Load learner and set search space in one go.
1. Run hyperparameter tuning with random search.
1. Get best performing hyperparameter configuration.

```{r, eval=FALSE}
library(mlr3tuningspaces)

task = tsk("penguins")

learner = lts(lrn("classif.rpart"))

instance = tune(
  method = "random_search",
  task = task,
  learner = learner,
  resampling = rsmp("cv", folds = 3),
  measure = msr("classif.acc"),
  term_evals = 10,
  batch_size = 5
)

instance$result
```

### Pipelines Example

Create a stacked learner:

1. Load task.
1. Choose list of base learners and super learner.
1. Convert to graph learner.
1. Train on german credit task.

```{r, eval=FALSE}
library(mlr3verse)
library(mlr3pipelines)

task = tsk("german_credit")

base_learners = list(
  lrn("classif.rpart", predict_type = "prob"),
  lrn("classif.kknn", predict_type = "prob")
)
super_learner = lrn("classif.log_reg")

graph_stack = pipeline_stacking(base_learners, super_learner)
graph_learner = as_learner(graph_stack)

graph_learner$train(task)
```

## Citing mlr3

If you use `mlr3`, please cite our JOSS article:

<div class="citation">
```
@Article{mlr3,
  title = {{mlr3}: A modern object-oriented machine learning framework in {R}},
  author = {Michel Lang and Martin Binder and Jakob Richter and Patrick Schratz and Florian Pfisterer and Stefan Coors and Quay Au and Giuseppe Casalicchio and Lars Kotthoff and Bernd Bischl},
  journal = {Journal of Open Source Software},
  year = {2019},
  month = {dec},
  doi = {10.21105/joss.01903},
  url = {https://joss.theoj.org/papers/10.21105/joss.01903},
}
```
</div>
