---
title: "Learners"
description: |
  Learners of the mlr3 ecosystem.
preview: images/logo_color.png
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(mlr3website)
library(htmltools)
library(mlr3misc)
library(data.table)
```

To keep the dependencies on other packages reasonable, the base package mlr3 only ships with with regression and classification trees from the `r cran_pkg("rpart")` package and some learners for debugging.
A subjective selection of implementations for essential ML algorithms can be found in `r mlr_pkg("mlr3learners")` package.
Survival learners are provided by `r mlr_pkg("mlr3proba")`, cluster learners via `r mlr_pkg("mlr3cluster")`.
Additional learners, including some learners which are not yet to be considered stable or which are not available on CRAN, are connected via the `r mlr_pkg("mlr3extralearners")` package.


```{r}
content = mlr3extralearners::list_mlr3learners()

# add link to name and package
content[, name := pmap_chr(list(name, mlr3_package, id), function(name, mlr3_package, id) as.character(a(href = sprintf("https://%s.mlr-org.com/reference/mlr_learners_%s.html", mlr3_package, id), name)))]

content$required_packages = link_pkgs(content$required_packages)

# add line breaks to cells
sdcols = c("properties", "feature_types", "predict_types")
content[, (sdcols) := map(.SD, function(col) map_chr(col, function(x) paste0(x, collapse = "\n"))), .SDcols = sdcols]

setkeyv(content, "class")
col_names = col.names = c("Name", "Class", "mlr3 Package", "Packages", "Properties", "Feature Types", "Predict Types")
```

## Classification

For classification tasks, the goal is to predict a label with only relatively few distinct values.
The following learners are available for for classification problems.

```{r, layout="l-body-outset"}
table_responsive(content["classif", !"class"], escape = FALSE, col.names = col_names)
```

## Regression

For regression tasks, the objective is predict a continuous numeric quantity.
The following learners are available for for regression problems.

```{r, layout="l-body-outset"}
table_responsive(content["regr", !"class", on = "class"], escape = FALSE, col.names = col_names)
```

## Cluster

In cluster analysis the goal is to group data into clusters, where each cluster contains similar observations.
The following learners are available for cluster problems.

```{r, layout="l-body-outset"}
table_responsive(content["clust", !"class", on = "class"], escape = FALSE, col.names = col_names)
```

## Density

In density estimation the learning task is to find the unknown distribution from which an i.i.d. data set is generated.
The following learners are available for density estimation problems.

```{r, layout="l-body-outset"}
table_responsive(content["dens", !"class", on = "class"], escape = FALSE, col.names = col_names)
```

## Survival

In survival analysis the aim is to predict the survival distribution of a given individual.
The following learners are available for survival problems.

```{r, layout="l-body-outset"}
table_responsive(content["surv", !"class", on = "class"], escape = FALSE, col.names = col_names)
```
