---
title: "Learners"
description: |
  Learners of the mlr3 ecosystem.
preview: images/logo_color.png
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(mlr3website)
library(htmltools)
library(mlr3misc)
library(data.table)
library(DT)
library(mlr3)
library(mlr3learners)
library(mlr3extralearners)

content = as.data.table(mlr_learners, objects = TRUE)
content[, base_package := map(object, function(x) strsplit(x$man, "::", TRUE)[[1]][1])]
content[, packages := pmap(list(packages, base_package), function(x, y) setdiff(x, c(y, "mlr3")))]

# add links
content[, key := pmap_chr(list(base_package, key), function(base_package, key) as.character(a(href = sprintf("https://%s.mlr-org.com/reference/mlr_learners_%s.html", base_package, key), key)))]
content[, packages := link_pkgs(packages)]

# add line breaks to cells
sdcols = c("packages", "properties", "feature_types", "predict_types")
content[, (sdcols) := map(.SD, function(col) map_chr(col, function(x) paste0(x, collapse = "<br/>"))), .SDcols = sdcols]

setcolorder(content, c("key", "label", "base_package", "packages", "feature_types", "predict_types", "properties"))
content[, `:=`(object = NULL, task_type = NULL)]
setnames(content,
  c("label", "key", "base_package", "packages", "feature_types", "predict_types", "properties"),
  c("Label", "Key", "Base Package", "Packages", "Feature Types", "Predict Types", "Properties"))
```

To keep the dependencies on other packages reasonable, the base package mlr3 only ships with with regression and classification trees from the `r cran_pkg("rpart")` package and some learners for debugging.
A subjective selection of implementations for essential ML algorithms can be found in `r mlr_pkg("mlr3learners")` package.
Survival learners are provided by `r mlr_pkg("mlr3proba")`, cluster learners via `r mlr_pkg("mlr3cluster")`.
Additional learners, including some learners which are not yet to be considered stable or which are not available on CRAN, are connected via the `r mlr_pkg("mlr3extralearners")` package.

```{r, layout="l-page"}
datatable(
  content,
  style = "bootstrap",
  escape = FALSE,
  options = list(paging = FALSE, scrollY = "60vh", scrollX = TRUE, info = FALSE),
  rownames= FALSE)
```
