---
title: "Learn mlr3"
description: |
  Useful resources for learning mlr3
preview: images/logo_color.png
site: distill::distill_website
bibliography: publications.bib
---

```{r setup, include=FALSE}
library(mlr3website)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include = FALSE}
# Collection of resources
# Resources are rendered in a 3 columns grid.
# Text description should be one sentence

library(mlr3misc)
library(htmltools)
library(data.table)

resource_grid = function(content) {
  div(class = "grid-container-package",
    pmap(content, function(category, name, icon, url, text) {
        div(class = "grid-item-package",
          h3(
            tags$i(class = sprintf("bi-%s", icon)),
            HTML("&nbsp;"),
            name
          ),
          paste0(text, "."),
          a(href = url, span(class = "grid-package-link"))
        )
    })
  )
}

content = rowwise_table(
    ~category,    ~name,               ~icon,       ~url,                                                    ~text,
    "general",    "Book",              "book",            "https://mlr3book.mlr-org.com/",                         "Central entry point to the mlr3verse",
    "general",    "Gallery",           "image",           "https://mlr3gallery.mlr-org.com/",                      "Collection of case studies and demos",
    "general",    "Reference Manuals", "question-circle", "packages.html",                                         "Reference manuals (via package overview)",
    "general",    "R6 Introduction",   "dice-6",          "https://r6.r-lib.org/articles/Introduction.html",       "Introduction to R6 classes",
    "general",    "Future Package",    "stack",           "https://futureverse.org/",                              "Homepage of the parallelization framework 'future'",

    "cheatsheet", "mlr3",              "file-pdf",        "https://cheatsheets.mlr-org.com/mlr3.pdf",              "Core package cheat sheet",
    "cheatsheet", "mlr3tuning",        "file-pdf",        "https://cheatsheets.mlr-org.com/mlr3tuning.pdf",        "Tuning cheat sheet",
    "cheatsheet", "mlr3fselect",       "file-pdf",        "https://cheatsheets.mlr-org.com/mlr3fselect.pdf",       "Feature selection cheat sheet",
    "cheatsheet", "mlr3pipelines",     "file-pdf",        "https://cheatsheets.mlr-org.com/mlr3pipelines.pdf",     "Pipelines cheat sheet",

    "video",      "useR2019 talk",     "youtube",         "https://www.youtube.com/watch?v=wsP2hiFnDQs",           "Short intro to mlr3",
    "video",      "useR2019 talk",     "youtube",         "https://www.youtube.com/watch?v=gEW5RxkbQuQ",           "Short intro to mlr3pipelines and mlr3tuning",
    "video",      "useR2020 tutorial", "youtube",         "https://www.youtube.com/watch?v=T43hO2o_nZw",           "Tutorial on mlr3, mlr3tuning and mlr3pipelines",
    "video",      "ODSC talk 2021",    "youtube",         "https://av.tib.eu/media/55271",                         "Into to mlr3spatiotempcv and mlr3spatial",

    "course",     "I2ML course",       "display",         "https://introduction-to-machine-learning.netlify.app/", "Introduction to ML course. Free video lectures, slides, quizzes. Exercises use mlr3",
    "course",     "mlr-outreach",      "display",         "https://github.com/mlr-org/mlr-outreach",               "Slides and other material for teaching mlr3"
)

setkeyv(content, "category")
```

## Documentation

Entry points to learn about mlr3.

```{r, echo = FALSE}
resource_grid(content[.("general")])
```


## Cheat Sheets

The essential things neatly summarized.
Perfectly printed out next to the keyboard or on a second monitor.

```{r, echo = FALSE}
resource_grid(content[.("cheatsheet")])
```

## Videos

Recorded tutorials and lectures we have given.

```{r, echo = FALSE}
resource_grid(content[.("video")])
```

## Courses/Lectures

Material from teaching at our universities.

```{r, echo = FALSE}
resource_grid(content[.("course")])
```


## Peer-reviewed Articles

A more scientific view on our the packages and the packages we depend on.

* @mlr3: about the base package `r mlr_pkg("mlr3")`
* @mlr3pipelines: building machine learning pipelines with `r mlr_pkg("mlr3pipelines")`
* @mlr3proba: probabilistic regression with `r mlr_pkg("mlr3proba")` (including survival analysis)
* @future: the parallelization framework package `r cran_pkg("future")` we build upon
* @checkmate: package `r cran_pkg("checkmate")` for argument checking and defensive programming
* @batchtools: parallelization framework `r cran_pkg("batchtools")` for high-performance computing clusters, used via `r cran_pkg("future")` or `r gh_pkg("mlr-org/mlr3batchmark")`
