---
sidebar: false
toc: false
---

# Learners

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(mlr3website)
library(htmltools)
library(mlr3misc)
library(data.table)
library(mlr3)
library(mlr3learners)
library(mlr3extralearners)

content = as.data.table(mlr_learners, objects = TRUE)
content[, base_package := map(object, function(x) strsplit(x$man, "::", TRUE)[[1]][1])]
content[, packages := pmap(list(packages, base_package), function(x, y) setdiff(x, c(y, "mlr3")))]

# add links
content[, key := pmap_chr(list(base_package, key), function(base_package, key) as.character(a(href = sprintf("https://%s.mlr-org.com/reference/mlr_learners_%s.html", base_package, key), key)))]
content[, packages := link_pkgs(packages, format = "html")]

# add line breaks to cells
sdcols = c("packages", "properties", "feature_types", "predict_types")
content[, (sdcols) := map(.SD, function(col) map_chr(col, function(x) paste0(x, collapse = "<br/>"))), .SDcols = sdcols]

setcolorder(content, c("key", "label", "base_package", "packages", "feature_types", "predict_types", "properties"))
content[, `:=`(object = NULL, task_type = NULL)]
setnames(content,
  c("label", "key", "base_package", "packages", "feature_types", "predict_types", "properties"),
  c("Label", "Key", "Base Package", "Packages", "Feature Types", "Predict Types", "Properties"))
```

To keep the dependencies on other packages reasonable, the base package mlr3 only ships with with regression and classification trees from the `r cran_pkg("rpart")` package and some learners for debugging.
A subjective selection of implementations for essential ML algorithms can be found in `r mlr_pkg("mlr3learners")` package.
Survival learners are provided by `r mlr_pkg("mlr3proba")`, cluster learners via `r mlr_pkg("mlr3cluster")`.
Additional learners, including some learners which are not yet to be considered stable or which are not available on CRAN, are connected via the `r mlr_pkg("mlr3extralearners")` package.

```{r}
#| column: page
DT::datatable(content,
  style = "bootstrap",
  escape = FALSE,
  options = list(paging = TRUE, scrollX = TRUE, info = FALSE),
  rownames = FALSE)
```

## Example Usage

Fit a `r ref("mlr_learners_classif.rpart", text = "classification tree")` on the `r ref("mlr_tasks_breast_cancer", text = "Wisconsin Breast Cancer Data Set")` and predict on left-out observations.

```{r, echo = TRUE}
library("mlr3verse")

# retrieve the task
task = tsk("breast_cancer")

# split into two partitions
split = partition(task)

# retrieve a learner
learner = lrn("classif.rpart", keep_model = TRUE, predict_type = "prob")

# fit decision tree
learner$train(task, split$train)

# access learned model
learner$model

# predict on data frame with new data
predictions = learner$predict_newdata(task$data(split$test))

# predict on subset of the task
predictions = learner$predict(task, split$test)

# inspect predictions
predictions
predictions$score(msr("classif.auc"))
autoplot(predictions, type = "roc")
```


