---
title: "Measures"
description: |
  Performance Measures of the mlr3 ecosystem.
preview: images/logo_color.png
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(mlr3website)
library(htmltools)
library(mlr3misc)
library(data.table)
requireNamespace("mlr3verse")

content = as.data.table(mlr3::mlr_measures)[, .(key, label, task_type, packages)]
content[, key := link_keys(key, prefix = "mlr_measures")]
content[, packages := link_pkgs(packages, remove = c("mlr3", "mlr3measures"))]
content[is.na(task_type), task_type := "generic"]

setnames(content, c("Key", "Label", "Task Type", "Packages"))
tabs = split(content, content[["Task Type"]], keep.by = FALSE)
```

Measures operate on `r ref("Prediction")` objects generated by learners.
They quantify the prediction by comparing prediction with ground truth.
The `r ref("Measure")` objects provide an abstraction for a plethora of performance measures.

## Classification

For classification tasks, the goal is to predict a label with only relatively few distinct values.
The following measures are available for for classification problems.

```{r, layout="l-body-outset"}
table_responsive(tabs$classif, escape = FALSE)
```

## Regression

For regression tasks, the objective is predict a continuous numeric quantity.
The following measures are available for for regression problems.

```{r, layout="l-body-outset"}
table_responsive(tabs$regr, escape = FALSE)
```

## Cluster

In cluster analysis the goal is to group data into clusters, where each cluster contains similar observations.
The following measures are available for cluster problems.

```{r, layout="l-body-outset"}
table_responsive(tabs$clust, escape = FALSE)
```
## Density


In density estimation the learning task is to find the unknown distribution from which an i.i.d. data set is generated.
The following measures are available for density estimation problems.

```{r, layout="l-body-outset"}
table_responsive(tabs$dens, escape = FALSE)
```

## Survival

In survival analysis the aim is to predict the survival distribution of a given individual.
The following measures are available for survival problems.

```{r, layout="l-body-outset"}
table_responsive(tabs$surv, escape = FALSE)
```

## Generic Measures

The following measures are independent of the task type.

```{r, layout="l-body-outset"}
table_responsive(tabs$generic, escape = FALSE)
```

