---
title: "Measures"
description: |
  Performance Measures of the mlr3 ecosystem.
preview: images/logo_color.png
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(mlr3website)
library(htmltools)
library(mlr3misc)
library(data.table)
requireNamespace("mlr3verse")

content = as.data.table(mlr3::mlr_measures)
content$packages = link_pkgs(content$packages, remove = c("mlr3", "mlr3measures"))

# add line breaks to cells
sdcols = c("task_properties")
content[, (sdcols) := map(.SD, function(col) map_chr(col, function(x) paste0(x, collapse = "\n"))), .SDcols = sdcols]

setkeyv(content, "task_type")
setnames(content,
    c("key", "packages", "predict_type", "task_properties"),
    c("Name", "Packages", "Predict Type", "Task Properties")
)
```


## Classification

For classification tasks, the goal is to predict a label with only relatively few distinct values.
The following measures are available for for classification problems.

```{r, layout="l-body-outset"}
table_responsive(content[.("classif"), !"task_type"], escape = FALSE)
```

## Regression

For regression tasks, the objective is predict a continuous numeric quantity.
The following measures are available for for regression problems.

```{r, layout="l-body-outset"}
table_responsive(content[.("regr"), !"task_type"], escape = FALSE)
```

## Cluster

In cluster analysis the goal is to group data into clusters, where each cluster contains similar observations.
The following measures are available for cluster problems.

```{r, layout="l-body-outset"}
table_responsive(content[.("clust"), !"task_type"], escape = FALSE)
```
## Density


In density estimation the learning task is to find the unknown distribution from which an i.i.d. data set is generated.
The following measures are available for density estimation problems.

```{r, layout="l-body-outset"}
table_responsive(content[.("dens"), !"task_type"], escape = FALSE)
```

## Survival

In survival analysis the aim is to predict the survival distribution of a given individual.
The following measures are available for survival problems.

```{r, layout="l-body-outset"}
table_responsive(content[.("surv"), !"task_type"], escape = FALSE)
```

## Generic Measures

The following measures are independent of the task type.

```{r, layout="l-body-outset"}
table_responsive(content[.(NA), !"task_type"], escape = FALSE)
```

