---
title: "Packages"
description: |
  Package collection of the mlr3 ecosystem.
preview: images/logo_color.png
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

The mlr3 ecosystem is a collection of R packages for machine learning.
The base package mlr3 only provides the basic building blocks for machine learning.
The extensions packages extent mlr3 with functionality for additional task types, learning algorithms, tuning algorithms, feature selection strategies, visualizations or preprocessing capabilities.
The packages are listed bellow with a short description.
For more information about the packages, check out their respective homepages.
The dot next to the package name indicates the [lifecycle stage](https://lifecycle.r-lib.org/articles/stages.html).
Packages with a green dot <i class="bi-circle-fill status-stable"></i> are stable.
Experimental packages are marked with an orange dot <i class="bi-circle-fill status-experimental"></i>.
Planned packages are marked with a red dot <i class="bi-circle-fill status-planned"></i>.

![Graph of Extension Packages](https://raw.githubusercontent.com/mlr-org/mlr3/master/man/figures/mlr3verse.svg?sanitize=true)

```{r, include = FALSE}
# Collection of mlr3 ecosystem packages
# Packages are rendered in a 3 columns grid.
# Text description should be one sentence

library(mlr3misc)
library(htmltools)
library(data.table)

package_grid = function(content) {
  div(class = "grid-container-package",
    pmap(content, function(category, name, status, url, text) {
        div(class = "grid-item-package",
          h3(
            a(href = url, name),
            tags$i(class = sprintf("bi-circle-fill status-%s", status))
          ),
          paste0(text, ".")
        )
    })
  )
}

content = rowwise_table(
    ~category,       ~name,               ~status,        ~url,                                         ~text,
    "core",          "mlr3",              "stable",       "https://mlr3.mlr-org.com/",                  "Basic building blocks for machine learning",
    "core",          "mlr3verse",         "stable",       "https://mlr3verse.mlr-org.com/",             "Meta-package intended to simplify both installation and loading of packages from the mlr3 ecosystem",
    "optimization",  "mlr3tuning",        "stable",       "https://mlr3tuning.mlr-org.com/",            "Hyperparameter tuning for mlr3 learners",
    "optimization",  "mlr3tuningspaces",  "stable",       "https://mlr3tuningspaces.mlr-org.com/",      "Collection of search spaces for hyperparameter tuning",
    "optimization",  "mlr3hyperband",     "stable",       "https://mlr3hyperband.mlr-org.com/",         "Successive halving and hyperband tuner for mlr3tuning",
    "optimization",  "mlr3mbo",           "experimental", "https://mlr3mbo.mlr-org.com/",               "Model-based optimization for mlr3tuning",
    "optimization",  "miesmuschel",       "experimental", "https://github.com/mlr-org/miesmuschel/",    "Flexible mixed integer evolutionary strategies",
    "optimization",  "mlr3automl",        "experimental", "https://github.com/mlr-org/mlr3automl/",     "Automated machine learning",
    "featsel",       "mlr3filters",       "stable",       "https://mlr3filters.mlr-org.com/",           "Filter Feature Selection",
    "featsel",       "mlr3fselect",       "stable",       "https://mlr3fselect.mlr-org.com/",           "Wrapper Feature Selection",
    "data",          "mlr3db",            "stable",       "https://mlr3db.mlr-org.com/",                "Data backend to transparently work with databases",
    "data",          "mlr3oml",           "stable",       "https://mlr3oml.mlr-org.com/",               "Connector to OpenML",
    "data",          "mlr3data",          "stable",       "https://mlr3data.mlr-org.com/",              "Data sets and tasks",
    "learners",      "mlr3learners",      "stable",       "https://mlr3learners.mlr-org.com/",          "Essential learners for mlr3, maintained by the mlr-org team",
    "learners",      "mlr3extralearners", "stable",       "https://mlr3extralearners.mlr-org.com/",     "Extra learners for mlr3, implemented by the community",
    "learners",      "mlr3keras",         "experimental", "https://github.com/mlr-org/mlr3keras/",      "Deep learning with Keras",
    "visualization", "mlr3viz",           "stable",       "https://mlr3viz.mlr-org.com/",               "Visualizations for tasks, predictions, resample results and benchmarks",
    "pipelines",     "mlr3pipelines",     "stable",       "https://mlr3pipelines.mlr-org.com/",         "Dataflow programming toolkit",
    "tasks",         "mlr3spatiotempcv",  "stable",       "https://mlr3spatiotempcv.mlr-org.com/",      "Spatiotemporal resampling and visualization methods",
    "tasks",         "mlr3cluster",       "stable",       "https://mlr3cluster.mlr-org.com/",           "Cluster analysis",
    "tasks",         "mlr3proba",         "stable",       "https://mlr3proba.mlr-org.com/",             "Probabilistic predictions",
    "tasks",         "mlr3ordinal",       "experimental", "https://mlr3ordinal.mlr-org.com/",           "Ordinal regression",
    "tasks",         "mlr3multioutput",   "experimental", "https://github.com/mlr-org/mlr3multioutput", "Storing and working with multi-output tasks",
    "tasks",         "mlr3temporal",      "experimental", "https://mlr3temporal.mlr-org.com/",          "Time series analysis",
    "tasks",         "mlr3spatial",       "stable",       "https://mlr3spatial.mlr-org.com/",           "Spatial data backends and prediction functions",
    "tasks",         "mlr3multilabel",    "planned",      "https://github.com/mlr-org/mlr3multilabel",  "Multi-Label Classification",
    "tasks",         "mlr3fda",           "planned",      "https://github.com/mlr-org/mlr3fda",         "Functional Data Analysis",
    "utils",         "bbotk",             "stable",       "https://bbotk.mlr-org.com/",                 "Black-box optimization toolkit",
    "utils",         "paradox",           "stable",       "https://paradox.mlr-org.com/",               "Universal parameter space description and tools",
    "utils",         "mlr3misc",          "stable",       "https://mlr3misc.mlr-org.com/",              "Miscellaneous helper functions for mlr3",
    "utils",         "mlr3measures",      "stable",       "https://mlr3measures.mlr-org.com/",          "Performance measures for supervised learning",
    "utils",         "mlr3benchmark",     "stable",       "https://mlr3benchmark.mlr-org.com/",         "Analysis and tools for benchmarking",
    "utils",         "mlr3fairness",      "experimental", "https://mlr3fairness.mlr-org.com/",          "Fairness in Machine Learning",
    "utils",         "mlr3batchmark",     "experimental", "https://github.com/mlr-org/mlr3batchmark/",  "Connector between mlr3 and batchtools"
)

# index by categoty
# sort by status: stable first, then experimental, then planned
content[, status := factor(status, levels = c("stable", "experimental", "planned"), ordered = TRUE)]

# sort by status, keep order of table
setkeyv(content, c("category", "status"))

# sort by name
# setkeyv(content, c("category", "name"))
```

## <i class="bi-house-heart"></i> Core

```{r}
package_grid(content[.("core")])
```


## <i class="bi-sliders"></i> Optimization

```{r}
package_grid(content[.("optimization")])
```

## <i class="bi-funnel"></i> Feature Selection

```{r}
package_grid(content[.("featsel")])
```

## <i class="bi-hdd"></i> Data

```{r}
package_grid(content[.("data")])
```

## <i class="bi-pen"></i> Learners

```{r}
package_grid(content[.("learners")])
```

## <i class="bi-bar-chart-line"></i> Visualization

```{r}
package_grid(content[.("visualization")])
```

## <i class="bi-share"></i> Pipelines

```{r}
package_grid(content[.("pipelines")])
```

## <i class="bi-list-task"></i> Tasks

```{r}
package_grid(content[.("tasks")])
```

## <i class="bi-balloon"></i> Utilities

```{r}
package_grid(content[.("utils")])
```
