---
title: "Resamplings"
description: |
  Resampling Strategies of the mlr3 ecosystem.
preview: images/logo_color.png
site: distill::distill_website
---

Resamplings split the observations multiple times into two sets: training and test.
The former is used to fit the model, the latter is used to evaluate the predictions.
The `r ref("Resampling")` objects provide an abstraction for this procedure while respecting stratification as well as grouping/blocking if this is required by the `r ref("Task")`.

If only a single split is required (i.e., a holdout split), the `r ref("partition()")` function provides a single split into training and test set.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(mlr3website)
library(htmltools)
library(mlr3misc)
library(data.table)
library(DT)
requireNamespace("mlr3verse")

content = as.data.table(mlr3::mlr_resamplings)[, .(key, label, params, iters)]
content[, key := link_keys(key, prefix = "mlr_resamplings")]
content[, params := map(params, paste0, collapse = ", ")]
setnames(content, c("Key", "Label", "Parameters", "Default Iterations"))
```

```{r, layout="l-page"}
datatable(
  content,
  style = "bootstrap",
  escape = FALSE,
  options = list(paging = FALSE, scrollY = "60vh", scrollX = TRUE, info = FALSE),
  rownames= FALSE)
```
