---
bibliography: publications.bib
page-layout: full
title: Resources
sidebar: false
toc: true
---

{{< include _setup.qmd >}}

```{r}
#| include: false

resource_grid = function(content) {
  div(class = "grid grid-three",
    pmap(content, function(category, name, icon, url, text) {
      div(class = "grid__item",
        h3(class = "grid__title",
          tags$i(class = sprintf("bi-%s grid__title-icon", icon)),
          HTML("&nbsp;"),
          name
        ),
        p(
        paste0(text, "."),
        a(href = url, "aria-label" = sprintf("Go to %s", name), span(class = "grid__link-box"))
        )
      )
    })
  )
}

# Text description should be one sentence
content = rowwise_table(
  ~category, ~name, ~icon, ~url, ~text,
  "general", "Book", "book", "https://github.com/mlr-org/mlr3book/", "Central entry point to the mlr3verse",
  "general", "Gallery", "image", "https://mlr3gallery.mlr-org.com/", "Collection of case studies and demos",
  "general", "R6 Introduction", "dice-6", "https://r6.r-lib.org/articles/Introduction.html", "The mlr3 ecosystem is build on R6 classes. The link gives an introduction to R6",
  "general", "Future Package", "stack", "https://futureverse.org/", "Link to the future framework that is used to parallelize functions in mlr3",
  "general", "Developer Information", "wikipedia", "https://github.com/mlr-org/mlr3/wiki", "Link to mlr3 developer wiki",
  "cheatsheet", "mlr3", "file-pdf", "https://cheatsheets.mlr-org.com/mlr3.pdf", "Core package cheat sheet",
  "cheatsheet", "mlr3tuning", "file-pdf", "https://cheatsheets.mlr-org.com/mlr3tuning.pdf", "Tuning cheat sheet",
  "cheatsheet", "mlr3fselect", "file-pdf", "https://cheatsheets.mlr-org.com/mlr3fselect.pdf", "Feature selection cheat sheet",
  "cheatsheet", "mlr3pipelines", "file-pdf", "https://cheatsheets.mlr-org.com/mlr3pipelines.pdf", "Pipelines cheat sheet",
  "video", "useR2019 talk", "youtube", "https://www.youtube.com/watch?v=wsP2hiFnDQs", "Short intro to mlr3",
  "video", "useR2019 talk", "youtube", "https://www.youtube.com/watch?v=gEW5RxkbQuQ", "Short intro to mlr3pipelines and mlr3tuning",
  "video", "useR2020 tutorial", "youtube", "https://www.youtube.com/watch?v=T43hO2o_nZw", "Tutorial on mlr3, mlr3tuning and mlr3pipelines",
  "video", "ODSC talk 2021", "youtube", "https://av.tib.eu/media/55271", "Into to mlr3spatiotempcv and mlr3spatial",
  "course", "I2ML course", "display", "https://slds-lmu.github.io/i2ml/", "Introduction to ML course. Free video lectures, slides, quizzes. Exercises use mlr3",
  "course", "mlr-outreach", "display", "https://github.com/mlr-org/mlr-outreach", "Slides and other material for teaching mlr3"
)
setkeyv(content, "category")
```


## Documentation

Entry points to learn about mlr3.

```{r, echo = FALSE}
resource_grid(content[.("general")])
```


## Cheat Sheets

The essential things neatly summarized.
Perfectly printed out next to the keyboard or on a second monitor.

```{r, echo = FALSE}
resource_grid(content[.("cheatsheet")])
```

## Videos

Recorded tutorials and lectures we have given.

```{r, echo = FALSE}
resource_grid(content[.("video")])
```

## Courses/Lectures

Material from teaching at our universities.

```{r, echo = FALSE}
resource_grid(content[.("course")])
```


## Peer-reviewed Articles

A more scientific view on our packages and the packages we depend on.

* @mlr3: about the base package `r ref_pkg("mlr3")`
* @mlr3pipelines: building machine learning pipelines with `r ref_pkg("mlr3pipelines")`
* @mlr3proba: probabilistic regression with `r ref_pkg("mlr3proba")` (including survival analysis)
* @future: the parallelization framework package `r ref_pkg("future")` we build upon
* @checkmate: package `r ref_pkg("checkmate")` for argument checking and defensive programming
* @batchtools: parallelization framework `r ref_pkg("batchtools")` for high-performance computing clusters, used via `r ref_pkg("future")` or `r ref_pkg("mlr3batchmark")`

## External Tutorials

* @pargent2023tutorial: An Introduction to Machine Learning for Psychologists in R
* @Zhao2024: Tutorial on survival modeling with applications to omics data. [Tutorial Website](https://ocbe-uio.github.io/survomics/survomics.html).
* Toby Hocking has written various [tutorials on `mlr3`](https://tdhock.github.io/blog/2025/mlr3-tutorials/), including a [comparison with other ML frameworks](https://tdhock.github.io/blog/2023/comparing-ml-frameworks/).
