---
sidebar: false
toc: false
---

{{< include _setup.qmd >}}

```{r}
#| include: false

library(mlr3tuning)
library(mlr3hyperband)

lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")
```

# Tuners

Popular black-box optimization techniques are implemented in the `r ref_pkg("bbotk")` package.
The corresponding connectors to for tuning hyperparameters of learners or pipelines reside as `r ref("Tuner")` objects in package `r ref_pkg("mlr3tuning")`.
Additionally, packages `r ref_pkg("mlr3hyperband")` and `r ref_pkg("mlr-org/mlr3mbo")` provide some modern and sophisticated approaches.

All tuners operator on box-constrained tuning spaces which have to be defined by the user.
Some popular spaces from literature are readily available as [tuning spaces](tuning_spaces.html).

```{r}
#| echo: false
#| column: page

content = as.data.table(mlr_tuners, objects = TRUE)[, .(key, label, packages, properties, param_classes, object)]
content[, properties := map(properties, function(x) setdiff(x, "dependencies"))]
content[, object := NULL]

reactable(content, columns = list(
  key = colDef(name = "Key",
    cell = function(value) {
      mlr3book::ref(sprintf("%s_%s", "mlr_tuners", value), text = value, format = "htmltools")
    }
  ),
  label = colDef(name = "Label"),
  packages = colDef(name = "Packages", cell = function(value) {
    value = setdiff(value, "bbotk")
    if (length(value)) {
      map(value, function(x) tags$li(class = "object-table__list", mlr3book::ref_pkg(x, pkg = x, format = "htmltools"), " "))
    }
  }),
  properties = colDef(name = "Properties", cell = function(value) {
    if (length(value)) {
      tags$ul(map(value, function(x) tags$li(class = "object-table__list", x)))
    }
  }),
  param_classes = colDef(name = "Classes", cell = function(value) {
    if (length(value)) {
      tags$ul(map(value, function(x) tags$li(class = "object-table__list", x)))
    }
  })
  ),
  searchable = TRUE,
  pagination = FALSE
)
```

## Example Usage

Tune the hyperparameters of a `r ref("mlr_learners_classif.rpart", text = "classification tree")` on the `r ref("mlr_tasks_penguins", text = "Palmer Penguins")` data set with `r ref("mlr_tuners_random_search", text = "random search")`.

```{r, echo=TRUE}
library(mlr3verse)

# retrieve task
task = tsk("penguins")

# load learner and set search space
learner = lrn("classif.rpart",
  cp = to_tune(1e-04, 1e-1, logscale = TRUE),
  minsplit = to_tune(2, 128, logscale = TRUE)
)

# load tuner and set batch size
tuner = tnr("random_search", batch_size = 10)

# hyperparameter tuning on the palmer penguins data set
instance = tune(
  method = tuner,
  task = task,
  learner = learner,
  resampling = rsmp("holdout"),
  measure = msr("classif.ce"),
  term_evals = 50
)

# best performing hyperparameter configuration
instance$result

# surface plot
autoplot(instance, type = "surface")

# fit final model on complete data set
learner$param_set$values = instance$result_learner_param_vals
learner$train(task)

print(learner)
```
