---
title: "Visualization of spatial cross-validation partitioning"
authors: ["Patrick Schratz"]
date: 2018-07-25
categories: ["R", "r-bloggers"]
tags: ["spatial", "resampling", "spatial-CV", "rstats", "performance estimation"]
description: "Visualizing spatial cross-validation in mlr"

---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library("mlr")
library("BBmisc")
library("ParamHelpers")
```

# Introduction

In July `mlr` got a new feature that extended the support for spatial data: The ability to visualize spatial partitions in cross-validation (CV) [9d4f3](https://github.com/mlr-org/mlr/commit/9d4f391d182f2a50ba532e91b59184adeb88ceb1).
When one uses the resampling descriptions "SpCV" or "SpRepCV" in `mlr`, the k-means clustering approach after Brenning (2005) is used to partition the dataset into equally sized, spatially disjoint subsets.
See also [this](https://www.r-spatial.org/r/2018/03/03/spatial-modeling-mlr.html) post on r-spatial.org and the [mlr vignette about spatial data](http://mlr.mlr-org.com/articles/handling_of_spatial_data.html) for more information.

# Visualization of partitions

When using random partitiong in a normal cross-validation, one is usually not interested in the random pattern of the datasets train/test split
However, for spatial data this information is important since it can help identifying problematic folds (ones that did not converge or showed a bad performance) and also proves that the k-means clustering algorithm did a good job on partitioning the dataset.

The new function to visualize these partitions in `mlr` is called `createSpatialResamplingPlots()`.
It uses `ggplot2` and its new `geom_sf()` function to create spatial plots based on the resampling indices of a `resample()` object.
In this post we will use the examples of the function to demonstrate its use.

First, we create a resampling description `rdesc` using spatial partitioning with five folds and repeat it 4 times.
This `rdesc` object is put into a `resample()` call together with our example task for spatial matters, `spatial.task`.
Finally, we use the `classif.qda` learner to have a quick model fit.

```{r, results='hide', message=FALSE}
library(mlr)
rdesc = makeResampleDesc("SpRepCV", folds = 5, reps = 4)
resamp = resample(makeLearner("classif.qda"), spatial.task, rdesc)
```

Now we can use `createSpatialResamplingPlots()` to automatically create one plot for each fold of the `resamp` object.
Usually we do not want to plot all repetitions of the CV.
We can restrict the number of repetitions in the argument `repetitions`.

Besides the required arguments `task` and `resample`, the user can specifiy the coordinate reference system that should be used for the plots.
Here it is important to set the correct EPSG number in argument `crs` to receive accurate spatial plots.
In the background, `geom_sf()` (more specifically `coords_sf()`) will transform the CRS on the fly to EPSG: 4326.
This is done because lat/lon reference systems are better for plotting as UTM coordinates usually clutter the axis.
However, if you insist on using UTM projection instead of WGS84 (EPSG: 4326) you can set the EPSG code of your choice in argument `datum`.

```{r}
plots = createSpatialResamplingPlots(spatial.task, resamp, crs = 32717,
  repetitions = 2, x.axis.breaks = c(-79.065, -79.085),
  y.axis.breaks = c(-3.970, -4))
```

To avoid overlapping axis breaks you might want to set the axis breaks on your own as we did here.

Now we got a list of 2L back from `createSpatialResamplingPlots()`.
In the first list we got all the plots, one for each fold.
Since we used two repetitions and five folds, we have a total of ten instances in it.

The second list consists of labels for each plot.
These are automatically created by `createSpatialResamplingPlots()` and can serve as titles later on.
Note that for now we just created the `ggplot` objects (stored in the first list of the `plots` object).
We still need to plot them!

Single `ggplot` objects can be plotted by just extracting a certain object from the list, e.g. `plots[[1]][[3]]`.
This would plot fold #3 of repetition one.

```{r, fig.align='center', warning=FALSE}
plots[[1]][[3]]
```
However usually we want to visualize all plots in a grid.
For this purpose we highly recommend to use the `cowplot` package and its function `plot_grid()`.

The returned objects of `createSpatialResamplingPlots()` are already tailored to be used with this function.
We just need to hand over the list of plots and (optional) the labels:

```{r, fig.align='center', fig.width=10}
cowplot::plot_grid(plotlist = plots[["Plots"]], ncol = 5, nrow = 2,
  labels = plots[["Labels"]])
```

# Multiple resample objects

`createSpatialResamplingPlots()` is quite powerful and can also take multiple `resample()` objects as inputs with the aim to compare both.
A typical use case is to visualize the differences between spatial and non-spatial partitioning.

To do so, we first create two `resample()` objects (one using "SpRepCV", one using "RepCV"):

```{r, results='hide', message=FALSE}
rdesc1 = makeResampleDesc("SpRepCV", folds = 5, reps = 4)
r1 = resample(makeLearner("classif.qda"), spatial.task, rdesc1)
rdesc2 = makeResampleDesc("RepCV", folds = 5, reps = 4)
r2 = resample(makeLearner("classif.qda"), spatial.task, rdesc2)
```

Now we can hand over both objects using a named list.
This way the list names will also directly be used as a prefix in the resulting plot labels.

```{r, fig.align='center', fig.width=10}
plots = createSpatialResamplingPlots(spatial.task,
  list("SpRepCV" = r1, "RepCV" = r2), crs = 32717, repetitions = 1,
  x.axis.breaks = c(-79.055, -79.085), y.axis.breaks = c(-3.975, -4))

cowplot::plot_grid(plotlist = plots[["Plots"]], ncol = 5, nrow = 2,
  labels = plots[["Labels"]])
```

# References

Brenning, A. (2012). Spatial cross-validation and bootstrap for the assessment of prediction rules in remote sensing: The R package sperrorest. In 2012 IEEE International Geoscience and Remote Sensing Symposium. IEEE. https://doi.org/10.1109/igarss.2012.6352393
